pirouette: the error BEAST2 makes in inferring a phylogeny

### 2019-01-24 article.tex
Revised the article, thanks @Giappo

%comments
\newcommand{\giovanni}[1]{\textcolor{blue}{\textbf{[GL: #1]}}}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment and inference model to create a
    posterior of jointly-estimated phylogenies and model parameter estimates.
    When a new macro-evolutionary speciation \giovanni{maybe "diversification" is a better word} model is developed,
    a good first step is to measure the error BEAST2 makes on a known
    phylogeny derived from the new speciation process \giovanni{I'd use "diversification mechanisms" in place of "speciation process"}, 
    with its current set of inference models. \\
  \textbf{2. }
    Here, we present a free, libre and open-source R package, \verb;pirouette;
    that assesses the inference error BEAST2 makes based on a known/true 
    phylogeny. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    As \verb;pirouette; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package. \\
\end{abstract}

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given/known/true phylogeny \giovanni{I'd use: "given phylogeny of which full information is provided"}. This involves multiple steps,
described in detail below.

\giovanni{suggestion: expressing the same concepts of the "in brief" section in points could be an easier experience for the reader}
\giovanni{start example}

The pipeline is summarized by the following steps:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given phylogenies to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.
\giovanni{end example}

In brief: from a given phylogeny, an alignment is simulated, from which a BEAST2
posterior is inferred, of which its phylogenies are compared with the given
phylogeny. The more the posterior phylogenies match the given phylogeny,
the lower the error. There is also the option to generate a 'twin tree',
that goes through the same pipeline.

The first step simulates a DNA alignment from the given phylogeny.
One can specify a DNA sequence
of any length at the root of the phylogeny, a DNA mutation rate, a
site (also: \giovanni{I'd use "i.e."} nucleotide substitution) model, 
a clock model, a random number generator (RNG) seed and a location
where the alignment is saved to. This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

The second step selects an inference model, using the alignment.
The user can specify \giovanni{whether} to use the generative model or select a best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in \giovanni{the} inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (also: \giovanni{I'd use "i.e."} speciation model), the user needs
to specify which tree prior is used in \giovanni{the} inference. 
When selecting a \giovanni{I'd use "the"} best
model, the alignment is used to find the inference model that has the
highest evidence (also: \giovanni{I'd use "i.e."} marginal likelihood) from a set of inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;babette;, that call the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; inference model selection is not available on Windows.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
phylogeny using an error statistic, which is the nLTT 
statistic \cite{janzen2015approximate}. Additionally, the user can specify the
proportion of posterior phylogenies to discard (also called: the burn-in \giovanni{as far as I know the burn-in is to discard the start of a chain because it's still too far from the convergence, see http://www.beast2.org/glossary/}).

An optional step is to generate a 'twin tree', that will be
analyzed in the same way as the true tree. The twin tree has the same
topology as the given phylogeny, yet its branch lengths follow the same
distribution as a Yule (pure-birth) or (constant rate) birth-death tree prior.

@@

\begin{lstlisting}[language=R, floatplacement=H]
usthis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}
\giovanni{is "usthis" supposed to be "devtools"?}
To start using \verb;pirouette;, load its functions in the global namespace first:

@@

\begin{lstlisting}[language=R, floatplacement=H]
install_beast2()
\end{lstlisting}

@@

A first research question that \verb;pirouette; answers is:
What is the error BEAST2 makes from a phylogeny using the same 
speciation \giovanni{I'd use: "diversification"} model as it was generated by?

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. To do so, we must specify the DNA root sequence
and a mutation rate. In this example, the DNA root sequence consists
out of four block of 250 nucleotides each, where the per-nucleotide
mutation rate is on average \giovanni{I don't think it's correct to say that's "on average". Usually when you have a rate $\lambda$ you suppose to use a Poisson process generated by this rate: $\frac{\lambda^k e^\lambda}{k!}$. $\lambda$ turns out to be also the expected number of occurrences} 0.1 mutations per unit time:

@@

% Note that this syntax does not work yet, see
% https://github.com/richelbilderbeek/pirouette/issues/36
\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = create_bd_tree_prior()
)
\end{lstlisting}

@@

Here we start with a tree \giovanni{generated} from an unknown speciation \giovanni{I'd use: "diversification"} model,
that has four taxa and a crown age of 5:

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll use the same parameters here:

@@

% Note that this syntax does not work yet, see
% https://github.com/richelbilderbeek/pirouette/issues/36
\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

@@

\giovanni{I noticed that "phylogeny", "alignment params" and "model select param" are needed for all the three research questions. Some times, though, you repeat yourself. In an article you want to be as concise as possible so I'd suggest to avoid repetitions. One solution could be to summarize all the parameter settings in a summary scheme}

@@

Here we start again with a tree from an unknown speciation \giovanni{I'd use: "diversification"} model,
that has four taxa and a crown age of 5:

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll use the same parameters here:

% Note that this syntax does not work yet, see
% https://github.com/richelbilderbeek/pirouette/issues/36
\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

Also here, the third and fourth step have sensible defaults.

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branch lengths follow
a birth-death distribution. Creating a twinning parameter is easy,
as it has sensible default settings:

>>>>

%comments
\newcommand{\giovanni}[1]{\textcolor{blue}{\textbf{[GL: #1]}}}
\newcommand{\richel}[1]{\textcolor{orange}{\textbf{[RB: #1]}}}

@@

  \textbf{1. }
    BEAST2 is a popular Bayesian phylogenetics software tool,
    that takes an alignment and inference model to create a
    posterior of jointly-estimated phylogenies and model parameter estimates.
    When a new macro-evolutionary diversification model is developed,
    a good first step is to measure the error BEAST2 makes on a known
    phylogeny derived from a new diversification mechanism, 
    with its current set of inference models. \\
  \textbf{2. }
    Here, we present a free, libre and open-source R package, \verb;pirouette;
    that assesses the inference error BEAST2 makes based on a known/true 
    phylogeny. \\
  \textbf{3. }
    We describe \verb;pirouette;'s usage and the biological scientific
    question it can answer, including full examples. \\
  \textbf{4. }
    As \verb;pirouette; is designed to be of high quality and extendable, 
    we conclude by describing the further development of the package. \\
\end{abstract}

@@

The goal of \verb;pirouette; is to measure the inference error BEAST2
makes from a given phylogeny of which full information is provided.
This involves multiple steps, described in detail below.

The pipeline is summarized by the following steps:
\begin{enumerate}
    \item from a given phylogeny an alignment is simulated;
    \item the alignment is used as BEAST2 input to infer a posterior distribution of phylogenies;
    \item posterior phylogenies are compared with the given phylogenies to estimate the error we make. The bigger the match, the lower the error;
\end{enumerate}
There is also the option to generate a 'twin tree',
that goes through the same pipeline.

The first step simulates a DNA alignment from the given phylogeny.
One can specify a DNA sequence
of any length at the root of the phylogeny, a DNA mutation rate, a
site (i.e. nucleotide substitution) model, 
a clock model, a random number generator (RNG) seed and a location
where the alignment is saved to. This step is relatively fast, but longer
DNA alignments will noticeably slow down the inference step.

The second step selects an inference model, using the alignment.
The user can specify whether to use the generative model or select a best model
from a set of inference models. 
When selecting the generative model,
the site and clock model used in the alignment simulation are used
in the inference. Because the given phylogeny (on which the alignment is based)
may have followed any tree prior (i.e. speciation model), the user needs
to specify which tree prior is used in the inference. 
When selecting for the best
model, the alignment is used to find the inference model that has the
highest evidence (i.e. marginal likelihood) from a set of inference models.
The evidence of an inference model is estimated using a nested sampling (NS)
approach, as described in \cite{maturana2018model}. The nested sampling is
performed by \verb;babette;, that call the 'NS' BEAST2 package. 
Using BEAST2 packages (in a scripted way) can only be done under Linux and Mac; 
inference model selection is not available on Windows.

@@

The fourth step measures the inference error, using the phylogenies in the
Bayesian posterior. These phylogenies are compared to the given
phylogeny using an error statistic, which is the nLTT 
statistic \cite{janzen2015approximate} by default,
but also an custom error statistic is supplied, based
on the absolute difference in the gamma statistic [Pybus & Harvy]. 
Additionally, the user can specify the
proportion of posterior phylogenies to 
discard (i.e. the burn-in), throwing away the first 10%
of all phylogenies by default. This burn-in is used to discard
the part of the MCMC chain that has not yet converged to a
representative part of state space.

An optional step is to generate a 'twin tree', that will be
analyzed in the same way as the true tree. The twin tree has the same
topology as the given phylogeny, yet its branch lengths follow the same
distribution as a Yule (pure-birth) or (constant rate) birth-death tree prior.
The use of such a twin tree, is to assess the minimum level of noise (i.e. 
error) BEAST2 makes when the input (the twin phylogeny) has an ideal branch
length distribution.

@@

\begin{lstlisting}[language=R, floatplacement=H]
usethis::install_github("richelbilderbeek/pirouette")
\end{lstlisting}
\giovanni{is "usethis" supposed to be "devtools"?}
\richel{"usethis" is the next "devtools", "devtools" will be deprecated in the future}
To start using \verb;pirouette;, load its functions in the global namespace first:

@@

\begin{lstlisting}[language=R, floatplacement=H]
install_beast2()
\end{lstlisting}
For the option to select inference models,
\verb;pirouette; uses the "NS" BEAST2 package [Maturana et al.].
It can be installed from within R, using:

\begin{lstlisting}[language=R, floatplacement=H]
install_beast2_pkg("NS")
\end{lstlisting}

@@

A first research question that \verb;pirouette; answers is:
What is the error BEAST2 makes from a phylogeny using the same 
diversification model as it was generated by?

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. To do so, we must specify the DNA root sequence
and a mutation rate. In this example, the DNA root sequence consists
out of four block of 250 nucleotides each, where the per-nucleotide
mutation rate is 0.1 mutations per unit time.

@@

\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_gen_model_select_param(
  alignment_params = alignment_params,
  tree_prior = create_bd_tree_prior()
)
\end{lstlisting}

@@

Here we start with a tree generated from an unknown 
diversification model, that has four taxa and a crown age of 5:

@@

The first step in \verb;pirouette; is to simulate a DNA alignment from the 
given phylogeny. We'll re-use the alignment parameters of the previous example.

@@

\begin{lstlisting}[language=R, floatplacement=H]
model_select_param <- create_best_model_select_param()
\end{lstlisting}

@@

@@

We use all the same settings as the previous model:
phylogeny, alignment parameters, inference model selection,
shared inference parameters and error measuring parameters.

This time, we are interested in creating a twin tree. A twin tree
has the same topology as the given tree, yet its branch lengths follow
an idealized birth-death distribution. Creating a twinning parameter is easy,
as it has sensible default settings:
